\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-09 chapters/contribution.tex
Comments François patients

\chapter*{Personal contribution}
Arne Van Den Kerchove conceptualized the works and performed the analyses in
this thesis, under supervision and in consultation with prof. Hakim Si-Mohammed,
prof. Marc Van Hulle and prof. François Cabestaing.
Dr. Arno Libert and Dr. Benjamin Wittevrongel consulted on the analyses performed in
\autoref{sec:stbf-struct}
Data collection was performed by Arne Van Den Kerchove for the dataset
presented in \autoref{sec:covert-align} and by Arne Van Den Kerchove and
Juliette Meunier for the dataset presented in \autoref{sec:patients}.
Participant recruitment at TRAINM was performed by Alixe Willemssens under
supervision of dr. Edward Schietekatte, at Fondation Partage et Vie by Marie de
Moura under supervision of prof. Etienne Allart and at UZ Leuven Neuromuscular reference
center under supervision of prof. Kristl Claeys and prof. Philip Van Damme.
Prof. Etienne allart provided clinical consultation.

>>>>

\chapter*{Personal contribution}
Arne Van Den Kerchove conceptualized the works and performed the analyses in
this thesis, under supervision and in consultation with prof. Hakim Si-Mohammed,
prof. Marc Van Hulle and prof. François Cabestaing.
Dr. Arno Libert and Dr. Benjamin Wittevrongel consulted on the analyses performed in
\cref{sec:stbf-struct}
Data collection was performed by Arne Van Den Kerchove for the dataset
presented in \cref{sec:covert-align} and by Arne Van Den Kerchove and
Juliette Meunier for the dataset presented in \cref{sec:patients}.
Participant recruitment at TRAINM was performed by Alixe Willemssens under
supervision of dr. Edward Schietekatte, at Fondation Partage et Vie by Marie de
Moura under supervision of prof. Etienne Allart and at the UZ Leuven Neuromuscular
Reference Center under supervision of prof. Kristl Claeys and prof. Philip Van Damme.
Prof. Etienne Allart provided clinical consultation.

### 2024-10-09 chapters/covert_align.tex
Comments François patients

\chapter{Compensating \acs{erp} jitter for gaze-independence}
\label{sec:covert-align}
\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}}

@@

Overall, we observed a 5.10\%pt. accuracy increase with \ac{wcble} over
\ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
These results compare to the performance gain in \cite{Zisk2022}.
They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
step-wise Linear Discriminant Analysis  (SWLDA) for 6 ALS patients, whose SWLDA
performance also suffered from jitter.
Note that interpretation of this comparison may be challenging due to differences in
interface design (number of targets, \ac{isi}), subject population,
\ac{eeg} recording procedure, and available training data.

\subsection{Gaze-independence through cross-condition transfer}
\label{sec:cross_results}
To further back our claim of gaze-independence in the case where eye motor
control cannot be assumed, we evaluate our proposed decoder in a transfer
learning setting between \ac{vsa} conditions.
While performing overt attention requires gaze redirection for each target
selection, performing covert or split \ac{vsa} continuously still requires
sustained gaze fixation, which might not be possible for some patients who
could benefit from such an application.
Studying the transfer between conditions simulates what happens
when the user performs different \ac{vsa} conditions throughout the
experimental session.
Furthermore, if our decoder performs well in transfer-learning settings,
it must capture some information about the \ac{erp} responses that is independent of
the \ac{vsa} condition, and hence does not depend on gaze redirection to perform
these conditions.
We introduce an additional setting of interest here, namely a combination of \ac{vsa}
conditions, which represents those cases where patients cannot
redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
target they attend to.
For BNCI2014-009, this is implemented as an equal mix of overt and
covert \ac{vsa}; for CVSA-ERP, the combined condition represents an equal mix of overt,
covert, and split \ac{vsa}, disregarding parameter $d$.

@@

\cref{fig:single_trial_roc_auc_diff} shows that \ac{wcble} significantly improves
covert \ac{vsa} decoding.
This is advantageous for the development of a class of \ac{erp}-\ac{bci} interfaces for
patients who prefer to rest their gaze on a fixation cross on the screen,
avoiding the effort of redirecting their eye gaze for every selection.
Furthermore, the performance gain over the first-stage classifier in split
\ac{vsa} ($d=3$) and between-\ac{vsa} condition transfer settings are promising for
patients with even less eye motor control who may experience involuntary eye
movements or fixation fatigue and hence cannot keep their gaze fixed throughout an
entire \ac{bci} operation session.
\ac{wcble} would allow them to operate a \ac{bci} comfortably while directing
their gaze to whichever portion of the screen they prefer, even when there is
another target present or this location varies during the
course of operation.
Although \ac{wcble} did not significantly improve overt \ac{vsa} single-trial decoding,
\cref{fig:block_evaluation} shows that this does not negatively impact target
selection accuracy.
While target selection accuracy also did not improve for split \ac{vsa}, the
increase in single-trial performance in split ($d=3$) shows that an iterative
alignment procedure has the potential to improve over \ac{cble} and its first-stage
classifier in this case as well.

@@

\cite{Zisk2022} addressed P3 jitter in ALS patients by augmenting the training data once
with time-shifted copies based on \ac{cble}-estimated jitter.
While we aim to train the first-stage classifier without the effects of jitter
in its parameters, they do the opposite by intentionally jittering the
training data.
As their focus was on ALS patients and inter-session stability, they
did not assess how their method interacts with visuospatial attention.
We achieved a similar performance gain with our jitter compensation method, but
argue that our method can cope with more granular latency differences, as
\cite{Zisk2022} augment the data with just one positive and negative time
shift.

@@

Secondly, performance might be improved by venturing beyond the classical
target/non-target binary classification problem.
Due to, for example, the perifoveal stimulus cruciform model~\cite{Vanegas2013}, covert and split \ac{vsa}
responses might differ based on the relative position in the field of view of
their related stimulus, which could be exploited in a multi-class classification
problem.
Similarly, explicitly taking into account the characteristics of the distractor \ac{erp}
response might have a beneficial effect.
Thirdly, results were obtained in an offline and within-session evaluation,
which does not reflect true \ac{bci} operation.
Using multiple sessions with online feedback, the user could optimize their
performance over time by controlling attention or gaze.
Finally, since this work was conducted with patient applications in mind, we
should highlight that the gaze of participants in the conducted experiments was
cued and fixed, which is per definition impossible for the patients we consider.
We are currently conducting a patient study to further investigate whether the
studied \ac{vsa} conditions are appropriate and to what extent they occur in
natural patient \ac{bci} operation.

@@

Our aim was to improve gaze-independent \ac{bci} performance for spatially
organized visual event-related potential (\ac{erp}) paradigms by using a suitable
decoder.
Earlier results on \ac{bci} performance in covert visuospatial attention
(\ac{vsa}) performance prediction have shown that accounting for single-trial
latency jitter could improve gaze-independent decoding performance.
We applied Classifier-based Latency Estimation (\ac{cble}) as a decoder robust to latency
jitter, but found no increase in gaze-independent decoding performance.
To remedy this, we improved \ac{cble} by adapting it into \ac{cble} with Woody
iterations (\ac{wcble}), an iterative scheme using probabilistic latency estimation.
Results for \ac{wcble} within and across \ac{vsa} condition decoding show that
gaze-independent \ac{bci} performance can be improved at the decoding stage.
Overt decoding performance was not improved, but our proposed method can
provide added value for patients who are unable to operate a visual \ac{bci} in overt
attention mode.
Later studies should confirm whether our findings hold in patient
populations suffering from a variety of eye-motor impairments and develop a
solution that is capable of properly handling multiple non-time-locked \ac{erp}
components.

\clearpage
\begin{subappendices}
\section{\Acs{cble} algorithm}
\label{sec:wcble/app/cble-alg}
\begin{algorithm}[h]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N,
  \mathbf{l^\mathrm{train}}, \mathcal{C}(\cdot,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\State $\theta \gets
			\mathrm{train}_\mathcal{C}(\{\mat{X}_n^\mathrm{train}[:,s_1:s_2]\}_{n=1}^N,\mathbf{l^\mathrm{train}})$
      \algorithmiccomment{Train stage 1}
		\For{$n=1\ldots N$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\theta,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$
	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}
      \algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$

  \caption{\Acs{cble}}
	\label{alg:cble}
\end{algorithm}

\clearpage
\section{\Acs{wcble} algorithm}
\label{sec:wcble/app/wcble-alg}
\begin{algorithm}[h]
	\textsc{Train}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_n^\mathrm{train}\}_{n=1}^N, \mathbf{l^\mathrm{train}},
		\mathcal{C}(\cdot,f,\Pr), s_1, s_2$
	\begin{algorithmic}[1]
		\State $\mat{X}'_n \gets \mat{X}^\mathrm{train}_n \quad \forall \quad n=1\ldots
    N$
      \algorithmiccomment{Train stage 1}
		\Repeat
		\State $\theta^* \gets \mathrm{train}_\mathcal{C}(\{\mat{X}'_n[:,s_1:s_2]\}_0^{N-1},\mathbf{l^\mathrm{train}})$
		\For{$n=1\ldots N$}
    \State  $s_n \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n',\theta^*,l_n)\right]$
		\State $\mat{X}'_n \gets \mathrm{align}(\mat{X}^\mathrm{train}_n, s^*_n)$
		\EndFor
    \Until{convergence or maximum iterations reached}
    \For{$n=1\ldots N$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{n,s}^\mathrm{train} \gets
      f(\mat{X}_n^\mathrm{train}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{n,\mathrm{target}}^\mathrm{train} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\mathrm{target})\right]$
		    \State $s_{n,\mathrm{non-target}}^\mathrm{train}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_n^\mathrm{train},\theta^*,\textrm{non-target})\right]$
    \EndFor

	\end{algorithmic}
	\textbf{Output:} $\theta^*,
  \mat{Y}^\mathrm{train},
  \mathbf{s}_\mathrm{target}^\mathrm{train},
  \mathbf{s}_\mathrm{non-target}^\mathrm{train}$

	\smallskip \hrule \smallskip
	\textsc{Evaluate}
	\smallskip \hrule \smallskip
	\textbf{Input:} $\{\mat{X}_m^\mathrm{test}\}_{m=1}^N, \mathcal{C}(\theta^*,f, \Pr), s_1, s_2$
  \begin{algorithmic}[1]
		\For{$m=1\ldots M$}\algorithmiccomment{Feature extraction for stage 2}
		  \For{$s=1\ldots R$}
      \State $y_{m,s}^\mathrm{test} \gets
      f(\mat{X}_n^\mathrm{test}[:,s:s+(s_2-s_1)],\theta^*)$
        \EndFor
        \State $s_{m,\mathrm{target}}^\mathrm{test} \gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\mathrm{target})\right]$
		    \State $s_{m,\mathrm{non-target}}^\mathrm{test}\gets
			    \mathrm{median}\left[\Pr(s|\mat{X}_m^\mathrm{test},\theta^*,\textrm{non-target})\right]$
    \EndFor
	\end{algorithmic}
  \textbf{Output:} $\mat{Y}^\mathrm{test},
  \mathbf{s}_\mathrm{target}^\mathrm{test},
  \mathbf{s}_\mathrm{non-target}^\mathrm{test}$
  \caption{\Acs{wcble}}
	\label{alg:wcble}
\end{algorithm}
\end{subappendices}

>>>>

\chapter{Compensating \acs{erp} jitter for gaze-independence}
\label{sec:covert-align}
\emph{This chapter has been published as part of~\textcite{VanDenKerchove2024}.}

@@

Overall, we observed a 5.10\%pt. accuracy increase with \ac{wcble} over
\ac{tlda} for covert \ac{vsa} in the BNCI2014-009 dataset and 5.55\%pt. in the CVSA-ERP dataset.
These results compare to the performance gain in \cite{Zisk2022}.
They observed a 5.63\%pt. accuracy increase with 1-10 selection repetitions over
step-wise Linear Discriminant Analysis  (SWLDA) for 6 participants with
\ac{als}, whose SWLDA performance also suffered from jitter.
Note that interpretation of this comparison may be challenging due to differences in
interface design (number of targets, \ac{isi}), subject population,
\ac{eeg} recording procedure, and available training data.

\subsection{Gaze-independence through cross-condition transfer}
\label{sec:cross_results}
To further back our claim of gaze-independence in the case where eye motor
control cannot be assumed, we evaluate our proposed decoder in a transfer
learning setting between \ac{vsa} conditions.
While performing overt attention requires gaze redirection for each target
selection, performing covert or split \ac{vsa} continuously still requires
sustained gaze fixation, which might be impaired in those who
could benefit from such an application.
Studying the transfer between conditions simulates what happens
when the user performs different \ac{vsa} conditions throughout the
experimental session.
Furthermore, if our decoder performs well in transfer-learning settings,
it must capture some information about the \ac{erp} responses that is independent of
the \ac{vsa} condition, and hence does not depend on gaze redirection to perform
these conditions.
We introduce an additional setting of interest here, namely a combination of \ac{vsa}
conditions, which represents those cases where the user cannot
redirect their gaze and hence can be in any one of the \ac{vsa} conditions depending on the
target they attend to.
For BNCI2014-009, this is implemented as an equal mix of overt and
covert \ac{vsa}; for CVSA-ERP, the combined condition represents an equal mix of overt,
covert, and split \ac{vsa}, disregarding parameter $d$.

@@

\cref{fig:single_trial_roc_auc_diff} shows that \ac{wcble} significantly improves
covert \ac{vsa} decoding.
This is advantageous for the development of a class of \ac{erp}-\ac{bci} interfaces for
users who prefer to rest their gaze on a fixation cross on the screen,
avoiding the effort of redirecting their eye gaze for every selection.
Furthermore, the performance gain over the first-stage classifier in split
\ac{vsa} ($d=3$) and between-\ac{vsa} condition transfer settings are promising for
users with even less eye motor control who may experience involuntary eye
movements or fixation fatigue and hence cannot keep their gaze fixed throughout an
entire \ac{bci} operation session.
\ac{wcble} would allow them to operate a \ac{bci} comfortably while directing
their gaze to whichever portion of the screen they prefer, even when there is
another target present or this location varies during the
course of operation.
Although \ac{wcble} did not significantly improve overt \ac{vsa} single-trial decoding,
\cref{fig:block_evaluation} shows that this does not negatively impact target
selection accuracy.
While target selection accuracy also did not improve for split \ac{vsa}, the
increase in single-trial performance in split ($d=3$) shows that an iterative
alignment procedure has the potential to improve over \ac{cble} and its first-stage
classifier in this case as well.

@@

\cite{Zisk2022} addressed P3 jitter inparticipants \ac{als} by augmenting the
training data once with time-shifted copies based on \ac{cble}-estimated jitter.
While we aim to train the first-stage classifier without the effects of jitter
in its parameters, they do the opposite by intentionally jittering the
training data.
As their focus was on \ac{als} and inter-session stability, they
did not assess how their method interacts with visuospatial attention.
We achieved a similar performance gain with our jitter compensation method, but
argue that our method can cope with more granular latency differences, as
\cite{Zisk2022} augment the data with just one positive and negative time
shift.

@@

Secondly, performance might be improved by venturing beyond the classical
target/non-target binary classification problem.
Due to, for example, the perifoveal stimulus cruciform model~\cite{Vanegas2013}, covert and split \ac{vsa}
responses might differ based on the relative position in the field of view of
their related stimulus, which could be exploited in a multi-class classification
problem.
Similarly, explicitly taking into account the characteristics of the distractor \ac{erp}
response might have a beneficial effect.
Thirdly, results were obtained in an offline and within-session evaluation,
which does not reflect true \ac{bci} operation.
Using multiple sessions with online feedback, the user could optimize their
performance over time by controlling attention or gaze.
Finally, since this work was conducted with applications for individuals with
\ac{sspgi} in mind, we
should highlight that the gaze of participants in the conducted experiments was
cued and fixed, which is per definition impossible for the end user group we consider.
In \cref{sec:patients}, we present a study to further investigate whether the
studied \ac{vsa} conditions are appropriate and to what extent they occur when
individuals with \ac{sspgi} operate a \ac{bci}.

@@

Our aim was to improve gaze-independent \ac{bci} performance for spatially
organized visual event-related potential (\ac{erp}) paradigms by using a suitable
decoder.
Earlier results on \ac{bci} performance in covert visuospatial attention
(\ac{vsa}) performance prediction have shown that accounting for single-trial
latency jitter could improve gaze-independent decoding performance.
We applied Classifier-based Latency Estimation (\ac{cble}) as a decoder robust to latency
jitter, but found no increase in gaze-independent decoding performance.
To remedy this, we improved \ac{cble} by adapting it into \ac{cble} with Woody
iterations (\ac{wcble}), an iterative scheme using probabilistic latency estimation.
Results for \ac{wcble} within and across \ac{vsa} condition decoding show that
gaze-independent \ac{bci} performance can be improved at the decoding stage.
Overt decoding performance was not improved, but our proposed method can
provide added value for users who are unable to operate a visual \ac{bci} in overt
attention mode.
Later studies should confirm whether our findings hold in individuals with
severe physical impairment and a variety of eye-motor impairments, and develop a
solution that is capable of properly handling multiple non-time-locked \ac{erp}
components.

### 2024-10-09 chapters/gaze_independence.tex
Comments François patients

\subsection{Incidence of eye motor impairment}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups unfortunately also suffer
to some extent from eye motor impairment, requiring \acp{bci} adapted to their condition.
\Cref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus,
\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).},
other eye tremors
\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.},
gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.},
involuntary movements, impaired pursuit, \ldots), and even complete
ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only affects vision and coordination but also
their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  \centering
  \makebox[\textwidth][c]{%
  \input{chapters/gaze_independence_incidence.tex}
  }
  \caption[Incidence of eye motor impairment in selected \acs{bci} target
    patient populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} target
      patient populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

Among the most severely affected patients are stroke
patients~\cite{Pollock2011,Rowe2019}, mostly
due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
These patients might suffer from severe or complete eye motor impairment
from the onset of their condition.
Case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
with \ac{lis} due to stroke, the group with a complete lack of eye motor
control is very small.

In \ac{als}, a progressive disease affecting motor neurons, oculomotor issues
are also fairly common.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, studies show that minor issues are still
fairly common, especially for patients with bulbar onset
\ac{als}~\cite{Kang2018, Guo2022,Moss2012}.
Furthermore, \textcite{Hayashi1991} show that as these patients progress past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.

@@

\Ac{fa} is a neurodegenerative disease involving loss of coordinated movement.
This almost always affects eye
movements~\cite{Fahey2008,Hocking2010,Furman1983}, with various forms of
involuntary movements and trouble pursuing or fixating on targets.
They also gradually have more trouble speaking but often retain some muscular
control.

@@

\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only tetraplegic patients without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} patients with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late-stage ALS.}}

\newcommand\fnwolpawcrit{\footnote{
``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential \ac{bci} users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.''~\cite{Wolpaw2008}}}

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every patient has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing patients, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).

@@

The most severely impaired of these patients constitute the \ac{lis}
patient group, which forms one of the main \ac{bci} interest groups.
In this work, we use the term \ac{lis} for a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Hence, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.

@@

To summarize, for patients with eye motor impairment, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \acp{bci} hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
\acp{bci}.

@@

This specific target is embedded within a broader goal of enabling
effective communication for eye-motor impaired \ac{lis} patients.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high \ac{itr}.
Finally, our efforts also aim to improve \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in both gaze-dependent
and gaze-independent settings.

@@

After collecting data from healthy participants and developing initial decoders, we conducted
a patient study with gaze-impaired individuals to evaluate the impact of eye motor deficits
on \ac{bci} performance.
This study focused on individuals with \ac{sspgi}, recruited from neurorehabilitation centers
and specialized care homes, including participants with conditions such as \ac{als}, \ac{fa},
and stroke.

>>>>

\subsection{Incidence of eye motor impairment}
\label{sec:gaze-independent/gaze-dependence/incidence}
\Ac{sspi} is often caused by damage to the central or peripheral nervous
system, either through congenital diseases (\ac{cp}, \ac{fa}, \ldots),
neurodegenerative (\ac{als}, \ac{ms}, \ldots) or acquired (stroke and
\ac{tbi}).
Many individuals in these groups are unfortunately also afflicted by some form
of eye motor impairment, requiring \acp{bci} adapted to their condition.
\Cref{tab:incidence} reports the relatively high frequencies of
eye motor impairment, which can range from minor (nystagmus,
\footnote{Involuntary, rhythmic, and repetitive eye movements recognizable by their
consistent directionality (horizontal, vertical, or rotational).},
other eye tremors
\footnote{These can include square-wave jerks, saccadic intrusions, microtremors, or
microsaccades while resting or fixating the gaze.},
gaze fixation fatigue or discomfort, \ldots), to severe (partial ophthalmoplegia
\footnote{Weakness or limited paralysis of one or more of the muscles that control eye
movement, leading to restricted eye motion, but not complete paralysis.
Often, a specific movement direction (up-down, left-right) is preserved.},
involuntary movements, impaired pursuit, \ldots), and even complete
ophthalmoplegia\footnote{Full eye movement paralysis.} or eye motor paresis.
This not only affects vision and coordination but also
their ability to operate a visual \ac{bci}~\cite{FriedOken2020}.

\begin{table}
  \centering
  \makebox[\textwidth][c]{%
  \input{chapters/gaze_independence_incidence.tex}
  }
  \caption[Incidence of eye motor impairment in selected \acs{bci} user target
    populations.]{%
      Incidence of eye motor impairment in selected \ac{bci} user target
      populations. \acs{als}: \acl{als}, \acs{ms}: \acl{ms}, \acs{dmd}:
      \acl{dmd}, \acs{sma}: \acl{sma}, \acs{cp}: \acl{cp}, \acs{lis}: \acl{lis}.
      $+$: frequent, $-$: infrequent.
    }
    \label{tab:incidence}
\end{table}

Among the most affected are those recovering from stroke~\cite{Pollock2011,Rowe2019},
and most of all those with due to brainstem or cerebellar stroke~\cite{Moncayo2009,Bogousslavsky1987}.
Stroke can lead to the most severe severe or even complete eye motor impairment
from the onset of their condition, resulting in the~\ac{lis}.
However, case study reports~\cite{Patterson1986,Graber2016} show that even in individuals
with \ac{lis} due to stroke, the group with a complete lack of eye motor
control is very small.

\ac{als} is progressive disease affecting motor neurons.
This initially results in general weakness and loss of muscle tone, but
eventually leads to full body paralysis.
Although eye movement is often cited as one of the longest preserved
capabilities in \ac{als}, studies show that minor issues are still
fairly common~\cite{Kang2018, Guo2022,Moss2012}.
The bulbar-onset variant is characterized by an early loss of speech
and increased involvement of eye motor symptoms~\cite{Guo2022}.
Furthermore, \textcite{Hayashi1991} show that as \ac{als} progresses past
the point of independent breathing, symptoms will eventually also involve
eye muscle paralysis.
One of the goals of \ac{bci} has always been to support these individuals to
ensure quality of life.

@@

\Ac{fa} is a neurodegenerative disease affecting the
spinal cord, peripheral nervous system, and cerebellum, resulting in an
impairing loss of muscle coordination.
This almost always heavily affects eye
movements~\cite{Fahey2008,Hocking2010,Furman1983,Cook2017}, with various forms of
involuntary movements and trouble pursuing or fixating on targets.
They also gradually have more trouble speaking but often retain some muscular
control.

@@

\newcommand\fnlis{\footnote{Multiple definitions of \ac{lis} are encountered in
\ac{bci} and neurological literature.
Some definitions include only those with tetraplegy without eye movements used
for communications.
Others distinguish Complete Locked-in Syndrome (CLIS) with full body paralysis,
including no eye motor control at all, from a \ac{lis} state with some preserved eye
movements or minor motor output.
While some definitions only include stroke or \ac{tbi} with damage to
specific regions in the brain (midbrain, brainstem, or
cerebellum)~\cite{Smith2005}, it can also generally refer to the state of full body paralysis
or loss of muscle tone incurred in neurodegenerative diseases, combined with
the inability to speak, such as occurs in late-stage ALS.}}

\newcommand\fnwolpawcrit{\footnote{
``The first class consists of people who are truly totally locked-in (e.g.,
due to end-stage ALS or severe cerebral palsy), who have no remaining
useful neuromuscular control of any sort, including no eye movement.
\elide\ This class is very small. \elide\
The second class of potential \ac{bci} users comprises those who retain
a very limited capacity for neuromuscular control. This group includes
people who retain some useful eye movement or enough limb muscle
function to operate a single-switch system. Such control is often slow,
unreliable, or easily fatigued.
This group is much larger than the first.''~\cite{Wolpaw2006}}}

We believe it is generally not opportune to delve too deeply into the underlying
ophthalmological and neurological mechanisms for each of the specific
conditions, or even the exact symptoms.
In clinical reality, every \ac{bci} user has different symptoms, which lead to their own
unique set of preserved capabilities and visual skills.
From a solution-oriented \ac{bci} engineering point of view, the etiology of the symptoms can be abstracted away.
Instead of categorizing users by their etiology, we will refer to those who might benefit from
\ac{bci} assistive communication technology as having \ac{sspi}, in line with
the terminology used by~\textcite{FriedOken2020}.
If they have eye motor impairment that affects their use of visual \ac{bci} or
eye-tracking solutions, we use the term \ac{sspgi} (see \cref{fig:gaze_independence/venn}).

@@

The most severely impaired of the user groups above constitute the \ac{lis}
group, which forms one of the main \ac{bci} interest groups.
In this work, we use the term \ac{lis} for a situation of (near)
complete paralysis and difficulties or the inability to communicate independently,
even with assistive technology.
This corresponds to classes one and two defined by
\textcite{Wolpaw2006}\fnwolpawcrit.
Hence, some degree of severe or complete eye motor impairment is usually
necessary to qualify as locked-in\fnlis.

@@

To summarize, for individuals with \ac{sspgi}, gazing directly at a screen target may
be uncomfortable, impractical, or even impossible.
Hence, assistive devices that rely on eye tracking are often inefficient for
them.
Consequently, while \acp{bci} hold great promise for these individuals, conventional
gaze-dependent \ac{bci} solutions do not meet their needs due to the absence of gaze
control.
Therefore, the development of decoding strategies that account for covert
\ac{vsa} becomes crucial in the pursuit of high-performance gaze-independent
\acp{bci}.

@@

This specific target is embedded within a broader goal of enabling
effective communication for those with \ac{sspgi}.
Ultimately, we wish to design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities.
We believe this can be done with a non-invasive, spatially
organized visual \acp{bci} with high \ac{itr}.
Finally, our efforts also aim to improve \ac{erp} decoding performance
in general, which will also contribute to their effectiveness in both gaze-dependent
and gaze-independent settings.

@@

After collecting data from healthy participants and developing initial decoders, we conducted
a study with seven individuals with \ac{sspgi} to evaluate the impact of eye motor deficits
on \ac{bci} performance.
This study recruited participants from neurorehabilitation centers
and specialized care homes, including participants with conditions such as \ac{als}, \ac{fa},
and stroke.
